Comparing Parallel Sorting Algorithms on a High Performance Computing Cluster Using OpenMP

### 2020-11-16 paper/Paper.tex
Updated a lot

    \section{Introduction}
    Sorting is a fundamental problem to be solved in many algorithms and applications.
    Any algorithm or application, for instance, that depends on having some ordering to its data will likely deal with sorting in some form. Because of its pervasiveness, solving the sorting problem efficiently is highly desirable. For this reason, many efficient sorting algorithms have been developed. Three of the most popular are Mergesort, Quicksort, and Heapsort. All three of these algorithms have time complexities of $O(nlog(n))$ in the average case. Mergesort and Heapsort also achieve this time bound in the worst case, while Quicksort can in rare situations exhibit a runtime of $O(n^2)$ in the worst case. Since the asymptotic lower bound for comparison-based sorting algorithms such as these is $\Omega(nlog(n))$, significantly improving the performance of them isn't feasible with serial computation. For this reason, we must turn to parallel computation to achieve significant performance gains. In order to take advantage of the performance increases enabled by parallel computing, parallelized versions of various sorting algorithms have been developed and studied. While theoretical analysis of parallel sorting algorithms is useful for understanding the asymptotic differences in the runtimes, determining and analysing the empirical performance of the algorithms on a specific computer architecture is valuable for making an informed choice about which algorithm to choose for that particular architecture and under various conditions (e.g., the number of CPUs available and the size of the input array to sort).
    This study focuses on an empirical evaluation of parallelized versions of two of the most popular sorting algorithms: Quicksort and Mergesort. Heapsort is also analyzed as it is comparable to Quicksort and Mergesort, although it is not parallelized due to difficulties that we will discuss. The algorithms are parallelized with OpenMP so as to take advantage of multithreaded, shared-memory parallelism on Case Western Reserve's HPC Markov cluster.
    The layout of the paper is as follows: In Section II, we give a brief overview of parallel computing and sorting algorithms. Section III...

@@

    \subsection{Mergesort}
    Mergesort is a divide-and-conquer sorting algorithm that recursively divides an array into subarrays and merges them together in such a way that the original array is sorted. 
    The operation of this procedure on an array of eight elements is depicted in Figure \ref{mrg}.  
    \begin{figure}[h]
        \includegraphics[width=6cm]{merge.png} 
        \caption{Mergesort diagram from \cite{cormen_introduction_2009}}
        \label{mrg}
    \end{figure}
    Mergesort has a time complexity of $O(nlog(n))$ in both the average and worst case, although constant factors can make it worse than Quicksort in practice. 
    Given the divide-and-conquer paradigm that Mergesort uses, it is a natural candidate for parallelization. Mergesort follows this paradigm as follows: First the input array (of size $n$) is recursively divided in two until there are $n$ single-element subarrays (divide). Each single-element array is trivially sorted (conquer). From there, the sorted subarrays are merged together, resulting in a single array that is equivalent to the input array but in sorted order (combine). Given the mechanics of the algorithm, there are two primary ways to parallelize Mergesort: (1) Parallelize the divide stage (i.e., the two recursive calls to the Mergesort procedure) and (2) Parallelize the combine stage (i.e., the Merge procedure). It is possible to implement either (1), (2), or both. According to \cite{cormen_introduction_2009}, only parallelizing the recursive calls to Mergesort will result in diminishing returns on performance as the number of processors grows to more than a few dozen. In order to efficiently scale up to hundreds of processors, parallelizing the Merge procedure is necessary because it becomes the performance bottleneck.
    Our implementation of parallelized Mergesort is significantly based on \cite{radenski_mergesort}. The implementation parallelizes the recursive calls to Mergesort while implementing a serial Merge procedure. There are a couple of reasons for this design decision. First, the HPC architecture we are using (the Case Western Reserve Markov cluster) has well under one hundred CPU cores per node. Thus, parallel Mergesort with only the divide stage parallelized theoretically scales well within this core-count range. Second, parallelizing the Merge procedure proved to be difficult with OpenMP due to a lack of low-level control over threads.

@@

    \subsection{Heapsort}
    Heapsort is another $O(nlog(n))$ comparison-based sorting algorithm. 
    It, along with the heap data structure, was invented in 1964. \cite{forsythe_algorithms_1964}
    Heapsort first turns the dataset into a max heap, which is a binary tree where each parent node is greater then its children. 
    This process, called \textit{heapification}, is an $O(n)$ algorithum. 
    Sorting is performed by repeatably popping the root node (the maximum value) and re-heapifying. 
    This procedure takes advantage of the binary tree structure, and is worst case $O(n log (n))$ overall.  \cite{cormen_introduction_2009}
    Unfortunately, Heapsort is a poor candidate for paralleization, as it does not partition into subarrays and depends on the root node being the absolute maximum.

    \subsection{Quicksort}
    The final sort discussed here is Quicksort, another $O(n log(n))$ comparison sort. 
    Developed in 1961 \cite{hoare_algorithm_1961}, Quicksort is a partitioning sort that works by selecting a pivot element in an array and partitioning based on a comparison to the pivot, as shown in Figure \ref{qck}. 
    \begin{figure}[h]
        \includegraphics[width=6cm]{Quicksort.png} 
        \caption{Quicksort diagram. \href{https://www.techiedelight.com/quicksort/}{Source}}
        \label{qck}
    \end{figure}
    In particular, Quicksort is perhaps the most parallelizable sorting algorithm, and can achieve a linear speedup with few modifications. \cite{blelloch_programming_1996}
    This is because each subarray can be sorted independently, and this leads to speedup. 
    % TODO @Jason add detail on your implementation
    \section{Methodology}
	For our project, we used...
    For our project, we implemented our sorting algorithms in the C programming language. OpenMP \cite{openmp08} is used for achieving parallelization for our implementations of parallelized Mergesort and Quicksort.
    OpenMP facilitates multithreaded shared memory parallelism through the use of \textit{\#pragma} preprocessor directives in our C code.
	We ran our algorithms on Case Western Reserve University's HPC Markov cluster, which uses Intel Xeon x86\_64 Processors.

	The experiment consisted in running our sorting algorithms on array sizes of $10,000$, $100,000$, $1,000,000$, $10,000,000$, $100,000,000$, and $1,000,000,000$ while using CPU core numbers of $1$, $2$, $4$, $8$, and $16$ for each array size. The execution times for each run of each array size/CPU count
	combination were collected and recorded in CSV files. From there, we analyzed the data in order to compare the performance and scalability of our sorting algorithms, producing several graphs in order to provide visuals for aiding in the analysis. The execution of our algorithms on the Markov cluster was achieved through the use of the Simple Linux Utility for Resource Management \cite{yoo_slurm_2003} (SLURM), which enables repeatable, large scale experiments to be performed with precisely configured resources. We wrote SLURM batch scripts that execute our sorting algorithms on Markov cluster compute nodes for each combination of array size and CPU count and return the execution times in output files.
    % TODO do we need more detail?

@@

    This suggests that there is significant overhead with smaller array sizes, but this is overcome by efficeicny gains on larger array sizes. 
    \section{Conclusion}
    % TODO these are what we expect to see...
    As expected, the three $O(n log(n))$ algorithms perform better then Insertion sort. 
    Due to the more parallel nature of Quicksort and Mergesort, these algorithms benefit more from parallization then the more sequential Heapsort. 
    This is because the divide and conquer strategy is inherently more parallel, which should be taken into account when developing new algorithms to be run on parallel and distributed computing platforms. 
    % TODO further findings from implementation
    \bibliography{ref.bib}
    \bibliographystyle{ieeetran}
    \appendix
	All of our code and data can be found in our GitHub repo \href{https://github.com/causeyc/csds438_sorting}{here}. Our implementation of parallel Mergesort is located in mergesort/merge$\_$sort$\_$OpenMP.c. The parallel Quicksort implementation is in quicksort/quicksortOMP.c. Our serial Heapsort implementation is located in heapsort/heapsort.c. In order to run the experiment, clone the repo into a location on Case Western's HPC Markov cluster. From there, run the SLURM batch scripts. The first is \textbf{sorting.slurm} and is located in mergesort/. This script runs parallel Mergesort on a compute node on all combinations of array sizes and CPU core numbers (as explained in the main part of the paper). Once finished, an output SLURM file will be produced that contains the execution times of sorts. \textit{Mutatis mutandis} for Quicksort (the batch script is called \textbf{qs-job.slurm} and is located in quicksort/).
    % TODO slap the code in here
\end{document}

>>>>

    \section{Introduction}
    Sorting is a fundamental problem to be solved in many algorithms and applications.
    Any algorithm or application, for instance, that depends on having some ordering to its data will likely deal with sorting in some form. Because of its pervasiveness, solving the sorting problem efficiently is highly desirable. For this reason, many efficient sorting algorithms have been developed. Three of the most popular are Mergesort, Quicksort, and Heapsort. All three of these algorithms have time complexities of $O(nlog(n))$ in the average case. Mergesort and Heapsort also achieve this time bound in the worst case, while Quicksort can in rare situations exhibit a runtime of $O(n^2)$ in the worst case. Since the asymptotic lower bound for comparison-based sorting algorithms such as these is $\Omega(nlog(n))$, significantly improving the performance of them isn't feasible with serial computation. For this reason, we must turn to parallel computation to achieve significant performance gains. In order to take advantage of the performance increases enabled by parallel computing, parallelized versions of various sorting algorithms have been developed and studied. While theoretical analysis of parallel sorting algorithms is useful for understanding the asymptotic differences in the runtimes, determining and analysing the empirical performance of the algorithms on a specific computer architecture is valuable for making an informed choice about which algorithm to choose for that particular architecture and under various conditions (e.g., the number of CPUs available and the size of the input array to sort).
    This study focuses on an empirical evaluation of parallelized versions of two of the most popular sorting algorithms: Mergesort and Quicksort. Heapsort is also analyzed as it is comparable to Mergesort and Quicksort, although it is not parallelized due to difficulties that we will discuss. Instead, it is used as a reference serial algorithm to compare against our two parallelized algorithms. The algorithms are parallelized with OpenMP so as to take advantage of multithreaded, shared-memory parallelism on Case Western Reserve University's HPC Markov cluster. The criteria for evaluation are (i) execution time and (ii) scalability, i.e., the extent to which adding more processors decreases execution time.

\indent The layout of the paper is as follows: In Section II, we give an overview of parallel computing and the sorting algorithms that are included in our experiment and analysis. In Section III, we lay out our experimental methodology for comparing our sorting algorithms. Section IV presents our results and analysis. In Section V, we give our concluding remarks. Finally, in the Appendix, we provide information for how to access our code and data and how to run our experiment on the Case Western Markov cluster. % Weird space

@@

    \subsection{Mergesort}
    Mergesort is a divide-and-conquer sorting algorithm that recursively divides an array into subarrays and merges them together in such a way that the original array is sorted. 
    The operation of this procedure on an array of eight elements is depicted in Figure \ref{mrg}.  
    \begin{figure}[h]
        \includegraphics[width=6cm]{merge.png} 
        \caption{Mergesort diagram from \cite{cormen_introduction_2009}}
        \label{mrg}
    \end{figure}
    Mergesort has a time complexity of $O(nlog(n))$ in both the average and worst case, although constant factors can make it worse than Quicksort in practice. 
    Given the divide-and-conquer paradigm that Mergesort uses, it is a natural candidate for parallelization. Mergesort follows this paradigm as follows: First the input array (of size $n$) is recursively divided in two until there are $n$ single-element subarrays (divide). Each single-element array is trivially sorted (conquer). From there, the sorted subarrays are merged together, resulting in a single array that is equivalent to the input array but in sorted order (combine). Given the mechanics of the algorithm, there are two primary ways to parallelize Mergesort: (1) Parallelize the divide stage (i.e., the two recursive calls to the Mergesort procedure) and (2) Parallelize the combine stage (i.e., the Merge procedure). It is possible to implement either (1), (2), or both. According to \cite{cormen_introduction_2009}, only parallelizing the recursive calls to Mergesort will result in diminishing returns on performance as the number of processors grows to more than a few dozen. In order to efficiently scale up to hundreds of processors, parallelizing the Merge procedure is necessary because it becomes the performance bottleneck.
    Our implementation of parallelized Mergesort is significantly based on \cite{radenski_mergesort}. The implementation parallelizes the recursive calls to Mergesort while implementing a serial Merge procedure. There are a couple of reasons for this design decision. First, the HPC architecture we are using (the Case Western Reserve Markov cluster) has well under one hundred CPU cores per node. Thus, parallel Mergesort with only the divide stage parallelized theoretically scales well within this core-count range. Second, parallelizing the Merge procedure proved to be difficult with OpenMP due to a lack of low-level control over threads. OpenMP is used to parallelize Mergesort by making use of OpenMP's parallel sections construct. A parallel sections region is created, and each recursive call to Mergesort is enclosed in its own parallel section. This allows OpenMP to spawn multiple threads that can execute the recursive calls in parallel. Nested parallelism is used so that each level of the recursion can spawn more threads (if they are available). The performance gains from this parallelism are significant as will be demonstrated in our Results section.

@@

    \subsection{Heapsort}
    Heapsort is another $O(nlog(n))$ comparison-based sorting algorithm. 
    It, along with the heap data structure, was invented in 1964. \cite{forsythe_algorithms_1964}
    Heapsort first turns the dataset into a max heap, which is a binary tree where each parent node is greater then its children. 
    This process, called \textit{heapification}, is an $O(n)$ algorithum. 
    Sorting is performed by repeatably popping the root node (the maximum value) and re-heapifying. 
    This procedure takes advantage of the binary tree structure, and is worst case $O(n log (n))$ overall.  \cite{cormen_introduction_2009}
    Unfortunately, Heapsort is a poor candidate for parallelization, as it does not divide or partition the input array into subarrays and depends on the root node being the absolute maximum.

    \subsection{Quicksort}
    The final sort discussed here is Quicksort, yet another $O(nlog(n))$ comparison-based sorting algorithm.
    Developed in 1961 \cite{hoare_algorithm_1961}, Quicksort is a partitioning sort that works by selecting a pivot element in an array and partitioning based on a comparison to the pivot, as shown in Figure \ref{qck}. 
    \begin{figure}[h]
        \includegraphics[width=6cm]{Quicksort.png} 
        \caption{Quicksort diagram. \href{https://www.techiedelight.com/quicksort/}{Source}}
        \label{qck}
    \end{figure}
    Similarly to Mergesort, Quicksort follows the divide-and-conquer paradigm. This makes Quicksort a naturally parallelizable sorting algorithm that can in theory achieve a significant performance increase with few modifications to its core logic. \cite{blelloch_programming_1996}
    This is because each subarray can be sorted independently, and this leads to speedup on parallel computers.
    The fundamental steps of Quicksort are as follows: The first step is to partition the array around a selected “pivot” point to produce two subarrays. These subarrays are then recursively sorted until the entire array of elements are in the correct order. Like Mergesort, there are two ways to parallelize Quicksort.
	We can parallelize the recursive calls to the Quicksort procedure, we can parallelize the Partition procedure, or both. To keep the design comparable to our Mergesort implementation, we opted to simply parallelize the recursive calls to Quicksort. OpenMP is used to parallelize Quicksort by making use of
	OpenMP task directives. In OpenMP, the task directive can tell OpenMP to dedicate a
	thread to a specific block of code labeled by the task. In this case, when the two subarrays are selected, each recursive call to the Quicksort method gets a dedicated OpenMP task initiated to create a parallel region of threads to be called later. This increases efficiency incredibly when allocating more and more OpenMP threads for Quicksort to use as is demonstrated in our Results section.
    \section{Methodology}
    For our project and experiment, we implemented our sorting algorithms in the C programming language. OpenMP \cite{openmp08} is used for achieving parallelization for our implementations of parallelized Mergesort and Quicksort.
    OpenMP facilitates multithreaded shared memory parallelism through the use of \textit{\#pragma} preprocessor directives in our C code.
	We ran our algorithms on Case Western Reserve University's HPC Markov cluster, which uses Intel Xeon x86\_64 processors.

	The experiment consisted in running our sorting algorithms on array sizes of $10,000$, $100,000$, $1,000,000$, $10,000,000$, $100,000,000$, and $1,000,000,000$ while using CPU core numbers of $1$, $2$, $4$, $8$, and $16$ for each array size. The execution times for each run of each array size/CPU count
	combination were collected and recorded in CSV files. From there, we analyzed the data in order to compare the performance and scalability of our sorting algorithms, producing several graphs in order to provide visuals for aiding in the analysis. The execution of our algorithms on the Markov cluster was achieved through the use of the Simple Linux Utility for Resource Management \cite{yoo_slurm_2003} (SLURM), which enables repeatable, large scale experiments to be performed with precisely configured resources. We wrote SLURM batch scripts that execute our sorting algorithms on Markov cluster compute nodes for each combination of array size and CPU count, and the execution times (measured in seconds) are returned in output files. The primary comparison is between Mergesort and Quicksort; however, Heapsort is included in our results as a reference serial algorithm. Heapsort is used as a sorting algorithm in the C standard library. It thus serves as an example of the performance that can be expected from calling a standard library sorting function, and so it is useful to see how its performance compares to our parallelized Mergesort and Quicksort.
    % TODO do we need more detail?

@@

    This suggests that there is significant overhead with smaller array sizes, but this is overcome by efficeicny gains on larger array sizes. 
    \section{Conclusion}
    % TODO these are what we expect to see...
    As expected, the three $O(n log(n))$ algorithms perform better then Insertion sort. 
    Due to the more parallel nature of Quicksort and Mergesort, these algorithms benefit more from parallization then the more sequential Heapsort. 
    This is because the divide and conquer strategy is inherently more parallel, which should be taken into account when developing new algorithms to be run on parallel and distributed computing platforms. 
    % TODO further findings from implementation
    \bibliography{ref.bib}
    \bibliographystyle{ieeetran}
    \appendix
	All of our code and data can be found in our GitHub repo \href{https://github.com/causeyc/csds438_sorting}{here}. Our implementation of parallel Mergesort is located in mergesort/merge$\_$sort$\_$OpenMP.c. The parallel Quicksort implementation is in quicksort/quicksortOMP.c. Our serial Heapsort implementation is located in heapsort/heapsort.c. In order to run the experiment, clone the repo into a location on Case Western's HPC Markov cluster. From there, run the SLURM batch scripts. The first is \textbf{sorting.slurm} and is located in mergesort/. This script runs parallel Mergesort on a compute node on all combinations of array sizes and CPU core numbers (as explained in the main part of the paper). Once finished, an output SLURM file will be produced that contains the execution times of the sorts. \textit{Mutatis mutandis} for Quicksort (the batch script is called \textbf{qs-job.slurm} and is located in quicksort/). CSV files that we manually created from the data produced from our SLURM jobs are located in mergesort/ and quicksort/ and are called "MergeSort Data.csv" and "project.csv", respectively.
    % TODO slap the code in here
\end{document}
